Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-07-06 analysis/paper/paper.Rmd
making map for sampling area and revise

Taiwan had encountered multiple episodes of culture contact with China and Southeast Asia since Neolithic and became part of a global trading network during the early 17th century because of western European expansion that connected European and Asia. Compared to other places in Asia and Oceania, the impact of European colonialism on indigenous communities in East Asia appears to have been much less pronounced. Direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local indigenous communities remains unanswered. Indirect effects of colonialism is emphasized increasingly to discuss the colonial impact on the indigenous people in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, foreign trade goods introduced into local indigenous societies in a colonial or imperial contexts might cause significant transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Contrary to as passive receptors of imperial power, the agency of indigenous people in the colonial contexts has been broadly discussed to emphasize the ability of active adoptions of foreign effects, negotiation between colonized and colonizer, and resistance of indigenous people through daily cultural practices [@Dietler2015; @Given2004; @Rubertone2000; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005]. Studying the consumption of foreign goods is one way to understand indigenous experience of the colonial encounter and explore the agency of indigenous people by examining how those goods were distributed and used when obtained [@Dietler2005; @Mullins2011; @Scaramelli2005].

@@

In this paper, I explore the evidence from a site spans from late Iron Age to historical period by addressing the question of whether colonial influence on indigenous populations can be detected in ornaments recovered from the archaeological record at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, northern Taiwan had been involved in long-distance networks of East Asia since 14th century, and later encountered a large wave of Han Chinese migration in the 19th century [@Chen2005]. Northeastern Taiwan provides a good example to explore the indirect impacts on indigenous societies caused by comparing different episodes of culture contact based on a long-term perspective. My hypotheses are that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to the large scale exchange network, and the consumption of ornaments present some spatial patterns that indicate increasing social inequality. I also hypothesize that the use of ornaments declined during Chinese contact period.

# Cultural context of Ornaments in Northeastern Taiwan

Before European arrival, the native people had already established their own trading networks but the arrival of Europeans and their trade goods in the 17th century built on, but ...

Foreign ornaments including glass beads, stone beads, and metal objects first appeared in Taiwan around 1,800 BCE through the regional exchange network in Southeast Asia [@Wang2007]. Those new types of ornaments gradually replaced the earlier jade ornamental system in Taiwan during the transition from the Neolithic to the Iron Age. We observed a greater quantity and variety of foreign goods in the early 17th century during the European colonization period that involved the whole island in an international trade network on a large scale.

@Mullins2011---<They examine the ways gift-giving worked in the Middle Orinoco region of Venezuela, where missionaries hoped to secure indigenous alliances by providing goods for which they perceived a native demand. One class of goods missionaries provided was alcohol, which was used commonly in feasts and ritual contexts, but interjecting externally produced alcohol unseated existing production and consumption systems. Prior to colonization, drink had been produced by indigenous domestic units to enhance a family's prestige, but the introduction of imported spirits compelled hosts to obtain them through exchange, which forced increasing reliance on the production of cash crops and increased debt among indigenous consumers. This shift rendered drink “a commodity—a product that had to be bought, rather than produced,” so even though European alcohol was consumed within existing consumer contexts it led to dependency and exploitation--->

Although the northeastern Taiwan forms a circumscribed environment due to the surrounding mountains and sea, the sea provides a way through which indigenous people could interact with other groups from different regions, including other native indigenous groups in northern and southern Taiwan, foreign traders such as the Han people of China, and Europeans. This local exchange network of northeastern Taiwan had global connections after the 16th century because of its location on shipping routes between China and Japan [@Chen2005; @Wang2007]. The groups involved in the exchange network consisted by the Chinese traders, indigenous groups called Basay in northern Taiwan, and indigenous groups, Kavalan, in northeastern Taiwan. Historical records indicate that the Kavalan people were one of the Basay’s trading partners, and they exchanged rice, deer skin, and gold for metal tools or ornaments, beads, and Chinese ceramics [@Chen2005; @Hsieh2009; @LiandWu2006].

Historical records from Spanish and Dutch provide rich information about the historical and cultural background both for Europeans and indigenous people in the early 17th century. Spanish occupied and built forts in Northern Taiwan since 1624, from which Spanish missionaries often went to indigenous settlement and left behind numerous records from those places. The report of the Dominican Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) as a materials to exchange with necessities they need with other indigenous people. This trend then gradually spread to other areas, and even the Spanish soldiers used agate as bargaining chips for gambling, since agate equaled money [@LiandWu2006, pp. 132-149].  European contact was focused on trading with indigenous communities in northeastern Taiwan. Trading activity of many types of goode among these indigenous communities increased with the arrival of European colonisers. Thus, we predict that this trading activity resulted in an increase in the amount and diversity of ornaments coming into indigenous communities in northeastern Taiwan.

@@

Documents from the Qing dynasty usually describe the decorative purposes of the ornaments. In the Kavalan Subprefecture Gazetteer [@Chen2006[1852]: 308], the Kavalan Zhi Lue [@Ko1993[1837]: 11, 126) and the Dong Cha Ji Lue [@Yao1996[1829]: 77], it is mentioned that aborigines in Yilan used metal threads to weave so called golden carp-shaped ornaments. Their number was small and their value high, hence only the rich possessed them. Ordinary people wore agate beads or glass beads on their head or neck during various festivals. In 1895, at the beginning of Japanese rule over Taiwan, Ino Kanori [@Kanori1996, pp. 227-232] made field surveys among various ethnic groups in Taiwan. He describes the Plains Aborigines from Yilan as not using carp-shaped ornaments any more, but mentions that older people still use beads ornaments.

Recent ethnographic research of the Kavalan tribe describes agate beads as used in divination ceremonies, the subli, by female shamans [@Liu2008, pp. 133-134]. Interview records describe agate beads as valuable objects passed down from mother to daughter. When they were not used, they were often put on roof beams so that they were not seen by other people. But the origin of these beads is still not known. Later, because they were bought by Japanese and people in Taipei, agate beads became rarer and rarer [@Hsu1992, pp. 22] and carp-shaped ornaments had completely disappeared. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination. It was usually the female shamans that were responsible for divination rites, a technique passed down by ancestors. Most aborigines today don’t know the origin of agate beads any more, but their worth is still highlighted by oral history and through their scarcity.

@@

```{r, KWL-map, fig.cap = "Kiwulan location in Southeast Asia", out.width= "400px", echo=FALSE}
knitr::include_graphics("kiwulan-location-map.jpeg")
```

Kiwulan \@ref(fig:KWL-map) is located at Yilan, north of Lanyang Plain in northeastern Taiwan, surrounded by the Central Mountain Range in the west and the Pacific Ocean in the east. Kiwulan was excavated from 2001 to 2004, and the total excavation area was 3,814 m2. Most beads were found during the excavation and some of them through screens with 20 mm and 1.5 mm mesh. The archaeological evidence includes a rich amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures. The chronology of Kiwulan can be divided into Lower Layer Culture (700 - 1200 AD) and Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between based on a series of 32 radiocarbon dates[a], which coordinate to the late Iron Age and Proto-historical period in Taiwan. This paper focuses on the Upper Layer Culture which overlaps with the contact periods with the Europeans and Chinese. It provides a good examples to discuss the foreign impacts on local indigenous society through body ornaments that usually convey social and cultural meaning by displaying.

@@

# join two dataset to get coordinate
ornaments_period_join <-
  ornaments_period %>%
  mutate(period = as.factor(period)) %>% # seems need this line first 
  group_by(Pit_No, period) %>% 
  count(Categories) %>% 
  inner_join(AD_data, by = c("Pit_No" = "坑號"))
```

>>>>

Taiwan had been involved in regional exchange network of East Asia and Southeast Asia since Neolithic and became part of a global trading network during the early 17th century due to western European expansion that connected European and Asia. Compared to other places in Asia and Oceania, the impact of European colonialism on indigenous communities in East Asia appears to have been much less pronounced. Direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local indigenous communities remains unanswered. Indirect effects of colonialism is emphasized increasingly to discuss the colonial impact on the indigenous people in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, foreign trade goods introduced into local indigenous societies in a colonial or imperial contexts might cause significant transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Contrary to as passive receptors of imperial power, the agency of indigenous people in the colonial contexts has been broadly discussed to emphasize the ability of active adoptions of foreign effects, negotiation between colonized and colonizer, and resistance of indigenous people through daily cultural practices [@Dietler2015; @Given2004; @Rubertone2000; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005]. Studying the consumption of foreign goods is one way to understand indigenous experience of the colonial encounter and explore the agency of indigenous people by examining how those goods were distributed and used when obtained [@Dietler2005; @Mullins2011; @Scaramelli2005].

@@

In this paper, I explore the evidence from a site that spans the time from late Iron Age to historical period by addressing the question of whether colonial influence on indigenous populations can be detected in ornaments recovered from the archaeological record at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, northern Taiwan had been involved in long-distance networks of East Asia since 14th century, and later encountered a large wave of Han Chinese migration in the 19th century [@Chen2005]. As @Lape2003 points out based on case studies in island Southeast Asia, the study of multiple episodes of culture contact has potential to expand our understanding of the causes and effects of culture contact both locally and worldwide. Northeastern Taiwan provides a good example to explore the indirect impacts on indigenous societies by comparing archaeological evidence from different episodes of culture contact based on a long-term perspective. The variety and cultural context of trade ornaments will be reviewed, and their capacity to inform on how the distribution of ornaments reflect social changes will be examined. My hypothesis is that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to the large scale exchange network, and the consumption of ornaments present some spatial pattern that indicates increasing social inequality. I hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan.

# Cultural context of ornaments in northeastern Taiwan, Yilan

Before European arrival, the indigenous people in northern Taiwan had already established their own exchange network that defined as "inter-insular trade" by @Chen2005 [p. 12] to refer to small-scale regional trade on irregular basis . Despite being on the periphery of major trade ports at northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people since the prehistoric time, is a alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous people and another native group called Basay in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006]. This local exchange network of northeastern Taiwan had global connections after the 16th century because of its location on shipping routes between China and Japan, and was intensified in the 17th century when the arrival of Europeans and their trade goods [@Chen2005; @Wang2007].

Historical records from Spanish and Dutch provide rich information about the historical and cultural background both for Europeans and indigenous people in the early 17th century. Spanish occupied and built forts in Northern Taiwan since 1624, from which Spanish missionaries often went to indigenous settlement and left behind numerous records from those places. The report of the Dominican Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) as a materials to exchange with necessities they need with other indigenous people. This trend then gradually spread to other areas, and even the Spanish soldiers used agate as bargaining chips for gambling, since agate equaled money [@LiandWu2006, pp. 132-149].  European contact was focused on trading with indigenous communities in northeastern Taiwan. Trading activity of many types of goode among these indigenous communities increased with the arrival of European colonizers. Thus, we predict that this trading activity resulted in an increase in the amount and diversity of ornaments coming into indigenous communities in northeastern Taiwan.

@Mullins2011---<They examine the ways gift-giving worked in the Middle Orinoco region of Venezuela, where missionaries hoped to secure indigenous alliances by providing goods for which they perceived a native demand. One class of goods missionaries provided was alcohol, which was used commonly in feasts and ritual contexts, but interjecting externally produced alcohol unseated existing production and consumption systems. Prior to colonization, drink had been produced by indigenous domestic units to enhance a family's prestige, but the introduction of imported spirits compelled hosts to obtain them through exchange, which forced increasing reliance on the production of cash crops and increased debt among indigenous consumers. This shift rendered drink “a commodity—a product that had to be bought, rather than produced,” so even though European alcohol was consumed within existing consumer contexts it led to dependency and exploitation>

@@

Documents from the Qing dynasty usually describe the decorative purposes of the ornaments. In the Kavalan Subprefecture Gazetteer (Chen2006[1852]: 308), the Kavalan Zhi Lue (Ko1993[1837]: 11, 126) and the Dong Cha Ji Lue (Yao1996[1829]: 77), it is mentioned that aborigines in Yilan used metal threads to weave so called golden carp-shaped ornaments. Their number was small and their value high, hence only the rich possessed them. Ordinary people wore agate beads or glass beads on their head or neck during various festivals. In 1895, at the beginning of Japanese rule over Taiwan, Ino Kanori [@Kanori1996, pp. 227-232] made field surveys among various ethnic groups in Taiwan. He describes the Plains Aborigines from Yilan as not using carp-shaped ornaments any more, but mentions that older people still use beads ornaments.

Recent ethnographic research of the Kavalan tribe describes agate beads as used in divination ceremonies, the subli, by female shamans [@Liu2008, pp. 133-134]. Interview records describe agate beads as valuable objects passed down from mother to daughter. When they were not used, they were often put on roof beams so that they were not seen by other people. But the origin of these beads is still not known. Later, because they were bought by Japanese and people in Taipei, agate beads became rarer [@Hsu1992, pp. 22] and carp-shaped ornaments had completely disappeared. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination. It was usually the female shamans that were responsible for divination rites, a technique passed down by ancestors. Most aborigines today don’t know the origin of agate beads any more, but their worth is still highlighted by oral history and through their scarcity.

@@

```{r, KWL-map, fig.cap = "Kiwulan location in Southeast Asia", out.width= "400px", echo=FALSE}
knitr::include_graphics("kiwulan-location-map.png")
```

Kiwulan \@ref(fig:KWL-map) is located at northern Yilan and was excavated from 2001 to 2004 that creates 262 squares with the total area of 3,814 meter squares. Most beads were found in situ while digging and some were collected through screens with 20 mm and 1.5 mm mesh. The archaeological evidence includes a rich amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures. The chronology of Kiwulan can be divided into Lower Layer Culture (700 - 1200 AD) and Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between based on a series of 32 radiocarbon dates, which coordinate to the late Iron Age and Proto-historical period in Taiwan. This paper focuses on the Upper Layer Culture which overlaps with the contact periods with the Europeans and Chinese.

@@

# join two dataset to get coordinate
ornaments_period_join <-
  ornaments_period %>%
  mutate(period = as.factor(period)) %>% # seems need this line first 
  group_by(Pit_No, period) %>% 
  count(Categories) %>% 
  inner_join(AD_data, by = c("Pit_No" = "坑號"))

# mapping sampling area
ggplot(data = AD_data) +
  geom_sf(fill = NA) +
  geom_sf(data = sample_units, color = 'red', fill = NA) +
  geom_sf_text(aes(label = `坑號`), alpha = 0.7) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())
```
